Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-27 ranking-supermartingales.tex

Intuitively $Y_n$ is the rank of the program after $n$ steps of computation.
(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank, provided the term being reduced is not a value.

@@

\begin{theorem}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}, \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}]
\label{thm:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and set $T(s) := \min \set{n \mid Y_n(s) = 0}$.

@@

\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
However, it is true that
\[
\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
=
\int_{S} \mu(\dif s) \, f(M_n)(s)
\leq
\int_{S} \mu(\dif s) \, f(M_0)(s)
=
f(M)
\]
the inequality above follows from \Cref{lem:key rankable}.
}

@@

@@

\begin{corollary}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.

\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
\end{enumerate}
\end{corollary}

>>>>

Intuitively $Y_n$ gives the rank of the program after $n$ steps of computation.
(Assume that the rank of a term is 0 iff it is a value.) In a $\epsilon$-ranking supermartingale, each computation step causes a strict decrease in rank of at least $\epsilon$, provided the term in question is not a value.

@@

\begin{theorem}[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}, \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}]
\label{thm:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale, and define the r.v.~$T$ by $T(s) := \min \set{n \mid Y_n(s) = 0}$.

@@

\iffalse
\lo{We need to show that each $f(M_n)$ is integrable. LO claimed earlier that $0 \leq f(N) \leq f(M)$ for all $N \in \mathit{Rch}(M)$.
This is of false: take $f$ with $v \, \tsample \mapsto 1$ and $v \, \underline{r} \mapsto 2\, r$ for some value $v$. 
However, it is true that
\[
\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
=
\int_{S} \mu(\dif s) \, f(M_n)(s)
\leq
\int_{S} \mu(\dif s) \, f(M_0)(s)
=
f(M)
\]
the inequality above follows from \Cref{lem:key rankable}.
}
\fi

@@

As an immediate corollary of \Cref{lem:key rankable}, each $f(M_n)$ is integrable:
\[
\int_{S} \mu(\dif s) \, |f(M_n)(s)| 
=
\int_{S} \mu(\dif s) \, f(M_n)(s)
\leq
\int_{S} \mu(\dif s) \, f(M_0)(s)
=
f(M).
\]

@@

\begin{theorem}[Soundness]
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.

\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is PAST.
\end{enumerate}
\end{theorem}
